---
title: "BRCA_Mofa"
author: "Lukas Steininger"
date: "2025-06-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)

if (!requireNamespace("BiocManager", quietly = TRUE))
    #install.packages("BiocManager")
#BiocManager::install(version = "3.21")
#BiocManager::install("MOFA2")

library(MOFA2)
library(dplyr)

print(getwd())
```

# Load the data

Load the omics datasets, convert them into a matrix and transpose said matrix. MOFA expects the samples to be in the columns and the features to be in the rows.

```{r}
cp <- read.csv("../datasets/BRCA_Data/export_variance_threshold/copy_number.csv", row.names = 1)
cp_matix <-  t(as.matrix(cp))

gs <- read.csv("../datasets/BRCA_Data/export_variance_threshold/gene_expression.csv", row.names = 1)
gs_matrix <- t(as.matrix(gs))

pp <- read.csv("../datasets/BRCA_Data/export_variance_threshold/protein_levels.csv", row.names = 1)
pp_matrix <- t(as.matrix(pp))

mu <- read.csv("../datasets/BRCA_Data/export_variance_threshold/mutation.csv", row.names = 1)
mu_matrix <- t(as.matrix(mu))

data <- list(cp=t(as.matrix(cp)), gs=t(as.matrix(gs)), pp=t(as.matrix(pp)), mu=t(as.matrix(mu)))
```

# Load the target

The datasets contain multiple subtypes for each patient/row.

Included are:

-   vital.status [0, 1]

    -   alive and dead (?)

-   PR.status ['Positive' 'Negative', 'Performed but Not Available', 'Indeterminate', 'Not Performed']

    -   indicating whether the cancer cells show Progesterone receptors on their surface

-   ER.status ['Positive' 'Negative', 'Performed but Not Available', 'Indeterminate', 'Not Performed']

    -   indicating whether the cancer cells show Estrogen receptors on their surface

-   HER2.Final.Status ['Negative', 'Positive', 'Equivocal', 'Not Available']

    -   indicating whether the cancer cells overexpress the protein HER2 (Human Epidermal Growth Factor Receptor 2)

-   histological.type ['infiltrating ductal carcinoma', 'infiltrating lobular carcinoma']

    -   type of breast cancer

```{r}
# read targets of dataset
targets <- read.csv("../datasets/BRCA_Data/export_variance_threshold/targets.csv", row.names = 1)
```

# Multi-Omics Factor Analysis

## Define MOFA object

Define a MOFA object that holds the data and groups. This object is subsequently used to generate configurations for the actual analysis.

```{r}
# define location of model output
outfile = file.path(getwd(),"BRCA_output", "variance_threshold","model_BRCA_alive_dead.hdf5")

# create mofa object with groups
MOFAobject <- create_mofa(data, groups = targets$vital.status)

# add meta data
sample_metadata <- data.frame(
  sample = samples_metadata(MOFAobject)[1],
  group = samples_metadata(MOFAobject)[2],
  vital.status = targets$vital.status,
  ER.status = targets$ER.Status,
  PR.status = targets$PR.Status,
  HER2.status = targets$HER2.Final.Status,
  histological.type = targets$histological.type
)

samples_metadata(MOFAobject) <- sample_metadata

```
## Missingness per class and omics type

```{r}
plot_data_overview(MOFAobject)
```

## Generate configurations for analysis

```{r}
data_opts <- get_default_data_options(MOFAobject)
head(data_opts)
```

```{r}
model_opts <- get_default_model_options(MOFAobject)
model_opts$num_factors <- 5
model_opts$likelihoods <- c(
    cp = "gaussian",
    gs = "gaussian",
    pp = "gaussian",
    mu = "bernoulli"
)

head(model_opts)
```

```{r}
train_opts <- get_default_training_options(MOFAobject)
train_opts$drop_factor_threshold <- 0.1
head(train_opts)
```

## Perform MOFA

```{r}
MOFAobject_final <- prepare_mofa(
  object = MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)


MOFAobject.trained <- run_mofa(MOFAobject_final, outfile=outfile, use_basilisk = TRUE)
```
