---
title: "BRCA_Mofa"
author: "Lukas Steininger"
date: "2025-06-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("MOFA2")

library(MOFA2)
```

# Load the data

Load the omics datasets, convert them into a matrix and transpose said matrix. MOFA expects the samples to be in the columns and the features to be in the rows.

```{r}
cp <- read.csv("../datasets/BRCA_Data/split_variance_threshold/copy_number.csv", row.names = 1)
cp_matix <-  t(as.matrix(cp))

gs <- read.csv("../datasets/BRCA_Data/split_variance_threshold/gene_expression.csv", row.names = 1)
gs_matrix <- t(as.matrix(gs))

pp <- read.csv("../datasets/BRCA_Data/split_variance_threshold/protein_levels.csv", row.names = 1)
pp_matrix <- t(as.matrix(pp))

mu <- read.csv("../datasets/BRCA_Data/split_variance_threshold/mutation.csv", row.names = 1)
mu_matrix <- t(as.matrix(mu))

data <- list(cp=t(as.matrix(cp)), gs=t(as.matrix(gs)), pp=t(as.matrix(pp)))
```

# Load the target

The datasets contain multiple subtypes for each patient/row.

Included are:

-   vital.status [0, 1]

    -   alive and dead (?)

-   PR.status ['Positive' 'Negative', 'Performed but Not Available', 'Indeterminate', 'Not Performed']

    -   indicating whether the cancer cells show Progesterone receptors on their surface

-   ER.status ['Positive' 'Negative', 'Performed but Not Available', 'Indeterminate', 'Not Performed']

    -   indicating whether the cancer cells show Estrogen receptors on their surface

-   HER2.Final.Status ['Negative', 'Positive', 'Equivocal', 'Not Available']

    -   indicating whether the cancer cells overexpress the protein HER2 (Human Epidermal Growth Factor Receptor 2)

-   histological.type ['infiltrating ductal carcinoma', 'infiltrating lobular carcinoma']

    -   type of breast cancer

```{r}
targets <- read.csv("../datasets/BRCA_Data/split_variance_threshold/additional_info.csv")
targets <- subset(targets, select=-c(X))
```

# Multi-Omics Factor Analysis

## Define MOFA object

Define a MOFA object that holds the data and groups. This object is subsequently used to generate configurations for the actual analysis.

```{r}
MOFAobject <- create_mofa(data, groups = targets$ER.Status)
#MOFAobject <- create_mofa(data)

```

## Missingness per class and omics type

```{r}
plot_data_overview(MOFAobject)
```

## Generate configurations for analysis

```{r}
data_opts <- get_default_data_options(MOFAobject)
head(data_opts)
```

```{r}
model_opts <- get_default_model_options(MOFAobject)
model_opts$num_factors <- 5
head(model_opts)
```

```{r}
train_opts <- get_default_training_options(MOFAobject)
head(train_opts)
```

## Perform MOFA

```{r}
MOFAobject_final <- prepare_mofa(
  object = MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

outfile = file.path(getwd(),"BRCA_variance_cutoff","model_ER_Status.hdf5")
MOFAobject.trained <- run_mofa(MOFAobject_final, outfile=outfile, use_basilisk = TRUE)
```
