---
title: "STATegra_Mofa"
author: "Lukas Steininger"
date: "2025-06-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)

if (!requireNamespace("BiocManager", quietly = TRUE))
    #install.packages("BiocManager")
BiocManager::install("MOFA2")

library(MOFA2)
```

# Load the data

Load the omics datasets, convert them into a matrix and transpose said matrix. MOFA expects the samples to be in the columns and the features to be in the rows.

```{r}
proteomics <- read.csv("../datasets/STATegra/export_variance_threshold/proteomics.csv", row.names = 1)
proteomics_matix <-  t(as.matrix(proteomics))

metabolimics <- read.csv("../datasets/STATegra/export_variance_threshold/metabolomics.csv", row.names = 1)
metabolimics_matrix <- t(as.matrix(metabolimics))

transcriptomics <- read.csv("../datasets/STATegra/export_variance_threshold/transcriptomics.csv", row.names = 1)
transcriptomics_matrix <- t(as.matrix(transcriptomics))

data <- list(proteomics=proteomics_matix, metabolimics=metabolimics_matrix, transcriptomics=transcriptomics_matrix)
```

# Load the target

TBD

```{r}
targets <- read.csv("../datasets/STATegra/export_variance_threshold/targets.csv", row.names = 1)
```

# Multi-Omics Factor Analysis

## Define MOFA object

Define a MOFA object that holds the data and groups. This object is subsequently used to generate configurations for the actual analysis.

```{r}
# define location of model output
outfile = file.path(getwd(), "STATegra_output","variance_threshold","model_STATegra_con_ika.hdf5")

# create mofa object with groups
MOFAobject <- create_mofa(data, groups = targets$group)

sample_metadata <- data.frame(
  sample = samples_metadata(MOFAobject)[1],
  group = samples_metadata(MOFAobject)[2],
  experimental.group = targets$group,
  time = targets$time,
  time.classification = targets$time_class
)

samples_metadata(MOFAobject) <- sample_metadata
```

## Missingness per class and omics type

```{r}
plot_data_overview(MOFAobject)
```

## Generate configurations for analysis

```{r}
data_opts <- get_default_data_options(MOFAobject)
head(data_opts)
```

```{r}
model_opts <- get_default_model_options(MOFAobject)
model_opts$num_factors <- 3

head(model_opts)
```

```{r}
train_opts <- get_default_training_options(MOFAobject)

# Set parameters
train_opts$maxiter <- 1000

head(train_opts)
```

## Perform MOFA

```{r}
MOFAobject_final <- prepare_mofa(
  object = MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

MOFAobject.trained <- run_mofa(MOFAobject_final, outfile=outfile, use_basilisk = TRUE)
```
